---
title: "Vignette Title"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction 
Genomic Region Enrichment Analysis (GREA), is an R package which facilitates the ....
```{r}

```


## Example using simulated example with public data

We will be using using DESeq2 and biomaRt for differential expression analysis and subsequent annotation of genes. If you don't have these packages installed please run: 
```{r, eval=FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("DESeq2")
biocLite("biomaRt")
```


### Case study 
For this tutorial we will make use of a dataset comprised of 74 transcriptome profiles of CD4+ T cells from celiac disease patients and healthy controls, which have also been subjected to two different immune stimulation with CD3 and PMA [GSE69549](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE69549). We have pre-loaded this dataset into thre GREA package for this tutorial. 

```{r diffExpression, echo=FALSE, message=FALSE, warning=FALSE}
library(data.table)
library(tidyverse)
library(GREA)

data("col.data")
data("count.table")

pma_sampleNames <- rownames(col.data)[which(col.data$stimulation == "PMA")] 
pma_countTable <- count.table[,pma_sampleNames]
pma_colData <- col.data[pma_sampleNames,] #colData has sample names by row


library(DESeq2)
# setting "Control" samples as reference factor
pma_colData$type <- factor(pma_colData$type, levels= c("Control", "Coeliac"))
dds_pma <- DESeqDataSetFromMatrix(countData = pma_countTable,
                              colData = pma_colData,
                              design = ~ type)
dds_pma <- DESeq(dds_pma)
res_pma <- results(dds_pma)

## deffining differentially expressed genes for 
de_pma <- res_pma[which(res_pma$padj <= 0.01 & abs(res_pma$log2FoldChange ) >= 1),]

## number of differentially expressed genes. 
nrow(de_pma)
```

### Running GREA to test for enrichment of genomic regions within a gene set 
Now that we have a gene set defined by the differentially expressed genes between CD4+ T cells derived from healthy and celiac disease stimulated with PMA, we will run the `GREA.test` function to asseses if this is gene set is enriched genetic risk factors associated to autoimmune diseases.

#### Generating a random sets of genes 
Since genes are not uniformally distributed in the genome .... **** and we need to assess if the GWAS SNPs (which represent genomic regions) are enriched in neighboring genes. 
```{r binning gene.info, fig.align="center"}
# load the data.info object 
data("gene.info")

de_pma.ensmebl <- gene.info[match(rownames(de_pma), as.character(gene.info$gene_name)),"gene_id"]
# remove the NAs - symbols which could not be matched - usually non.coding genes and synonims 
de_pma.ensmebl <- de_pma.ensmebl[which(!is.na(de_pma.ensmebl))]

# we need to define the total number of genes which were tested, as this is the set of genes from which we will be able to generate the sampled gene sets. 
all.genes.ensembl <- gene.info[match(rownames(res_pma), as.character(gene.info$gene_name)),"gene_id"] 
all.genes.ensembl <- all.genes.ensembl[which(!is.na(all.genes.ensembl))]

# using the gene.info we will filtered out for the genes used in the anlaysis - and genes from autosomal chromosomes. 
gene.info.chr <- gene.info[which(gene.info$chr %in% c(1:22) == TRUE),]
gene.info.chr.filtered <- gene.info.chr[which(as.character(gene.info.chr$gene_id) %in% all.genes.ensembl),]

# with bin.genes.genomic.regions we will group all genes into bins which are defined by gene.binning.window
gene.binning.window <- 125000
gene.info.binned <- GREA::bin.genes.genomic.regions(gene.pos = gene.info.chr.filtered, window = gene.binning.window)

# we can also plot the distribution how many genes are in each of the bins we generated.

bin.nGenes <- table(gene.info.binned$bin.tag)
gene.info.binned$nGenes.in.bin <- as.numeric(bin.nGenes[as.character(gene.info.binned$bin.tag)])

ggplot(gene.info.binned, aes(x=nGenes.in.bin))+
  geom_bar(stat="count", position = "dodge", alpha=0.8)+
  theme_bw()+
  theme(text=element_text(size=10, family="Helvetica"))

```

As shown in the plot we can see that even thought that the plenty of bins contains only 1 gene (5,170), roughly 40% of the bins contains 2 or more genes. 

Now the object `gene.info.binned` can be used to generate random gene sets for this particular example. This might take a few minutes... 
```{r sampled.gene.sets}

de_pma.sampled.gene.sets <-  GREA::get.sampled.genes(input.genes = de_pma.ensmebl,
                                               gene.info.binned = gene.info.binned,
                                               binning.widow = 125000,
                                               n.genes.sets = 1000, 
                                               n.excess.genes.limit = 10,
                                               parallel = TRUE, 
                                               n.clusters = 3)


```

Once we have our random gene sets we can use them to create a null distribution and ascertain if our query gene set `get.sampled.genes`
```{r GREA.test, fig.align="center", fig.height=5, fig.width=7}
# loading genesets from Immuno-chip derived GWAS hits. 
data("ib.traits.genes.list")

de_pma.GREA <- GREA::GREA.test(query.gene.set = de_pma.ensmebl, 
                        sampled.genes.sets = de_pma.sampled.gene.sets, 
                        genomic.region.gene.sets = ib.traits.genes.list, 
                        plot = TRUE)
de_pma.GREA$GREA.summary

plot(de_pma.GREA$null.dist.plot)
```

Only 1 trait (PBC) seems to enriched with genes that are present in our 
